---
title: "DataCleaning"
author: "Gerardo Gutierrez"
date: "2025-11-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load necessary libraries
library(dplyr) # data wrangling
library(tidyr) # data tidying and reshaping
library(ggplot2) # pretty plots
library(readr) 
# Load the dataset
data <- read_csv("../data/raw_data.csv")
```
# Exploratory Data Analysis
```{r}
# Inspect the dataset
View(data)
head(data)


```

```{r}
summary(data)
```

```{r}
str(data)

```

```{r}
# Check for missing values
colSums(is.na(data))
```

```{r}
colnames(data)
```

```{r}
# plot correlation heatmap of numeric values

numeric_data <- data %>%
  select(where(is.numeric))

library(ggcorrplot)
cor_matrix <- round(cor(numeric_data, use = "complete.obs"), 2)
ggcorrplot(cor_matrix, lab = TRUE)
```

```{r}
# check unique values for all perfectly correlated variables
correlated_var <- c("Hectares", 'Seedrate(in Kg)', "LP_Mainfield(in Tonnes)", "Nursery area (Cents)", "LP_nurseryarea(in Tonnes)", "DAP_20days", "Urea_40Days", "Potassh_50Days", "Micronutrients_70Days", "Pest_60Day(in ml)", "Trash(in bundles)", "Weed28D_thiobencarb")

for (var in correlated_var) {
  cat("Unique values for", var, ":\n")
  print(unique(data[[var]]))
  cat("\n")
}
```


```{r}
# drop perfectlycorrelated variables except seedrate

# correlated vars = hectares,  seedrate (keep) lp mainfield, nursery area, lp nursery area in tonnes, dap20days, urea40days, potassh50days, micronutrients 70 days, pest 60days, trash, weed28d thiobencarb
drop_var <- c("Hectares", "LP_Mainfield(in Tonnes)", "Nursery area (Cents)", "LP_nurseryarea(in Tonnes)", "DAP_20days", "Urea_40Days", "Potassh_50Days", "Micronutrients_70Days", "Pest_60Day(in ml)", "Trash(in bundles)", "Weed28D_thiobencarb")

numeric_data_2 <- numeric_data %>%
  select(-all_of(drop_var))

# plot second heat map without correlated variables
cor_matrix_2 <- round(cor(numeric_data_2, use = "complete.obs"), 2)
ggcorrplot(cor_matrix_2, lab = TRUE)
```

```{r}
# convert categorical variables to factors
data <- data %>%
  mutate(across(where(~ is.character(.) || is.logical(.)), as.factor))
```

```{r}
# print categorical columns
categorical_cols <- data %>%
  select(where(is.factor)) %>%
  colnames()
print(categorical_cols)
```

```{r}
# print numeric data 2 columns
colnames(numeric_data_2)
```

```{r}
# concatenate numeric_data_2 with categorical data
categorical_data <- data %>%
  select(where(is.factor))
data_drop_cor<- cbind(numeric_data_2, categorical_data)
data_drop_cor

```

## Model 1
```{r}
# build model with data drop corr
model_1 <- lm(`Paddy yield(in Kg)` ~ ., data = data_drop_cor)
summary(model_1)
plot(model_1)
```
output from model 1 


Call:
lm(formula = `Paddy yield(in Kg)` ~ ., data = data_drop_cor)

Residuals:
    Min      1Q  Median      3Q     Max 
-4573.1  -466.2   103.9   649.1  1843.5 

Coefficients: (42 not defined because of singularities)
                                       Estimate Std. Error t value Pr(>|t|)    
(Intercept)                          -5960.0277 17184.7632  -0.347   0.7288    
`Seedrate(in Kg)`                      254.5612     0.4979 511.285   <2e-16 ***
`30DRain( in mm)`                      -87.8316   199.2548  -0.441   0.6594    
`30DAI(in mm)`                               NA         NA      NA       NA    
`30_50DRain( in mm)`                    35.0988   107.4816   0.327   0.7440    
`30_50DAI(in mm)`                            NA         NA      NA       NA    
`51_70DRain(in mm)`                          NA         NA      NA       NA    
`51_70AI(in mm)`                             NA         NA      NA       NA    
`71_105DRain(in mm)`                         NA         NA      NA       NA    
`71_105DAI(in mm)`                           NA         NA      NA       NA    
`Min temp_D1_D30`                      -33.6974    55.7843  -0.604   0.5458    
`Max temp_D1_D30`                      -14.5666    41.9808  -0.347   0.7286    
`Min temp_D31_D60`                      62.0278    40.4484   1.534   0.1253    
`Max temp_D31_D60`                           NA         NA      NA       NA    
`Min temp_D61_D90`                           NA         NA      NA       NA    
`Max temp_D61_D90`                           NA         NA      NA       NA    
`Min temp_D91_D120`                          NA         NA      NA       NA    
`Max temp_D91_D120`                          NA         NA      NA       NA    
`Inst Wind Speed_D1_D30(in Knots)`           NA         NA      NA       NA    
`Inst Wind Speed_D31_D60(in Knots)`          NA         NA      NA       NA    
`Inst Wind Speed_D61_D90(in Knots)`          NA         NA      NA       NA    
`Inst Wind Speed_D91_D120(in Knots)`         NA         NA      NA       NA    
`Relative Humidity_D1_D30`                   NA         NA      NA       NA    
`Relative Humidity_D31_D60`                  NA         NA      NA       NA    
`Relative Humidity_D61_D90`                  NA         NA      NA       NA    
`Relative Humidity_D91_D120`                 NA         NA      NA       NA    
AgriblockCuddalore                           NA         NA      NA       NA    
AgriblockKallakurichi                        NA         NA      NA       NA    
AgriblockKurinjipadi                         NA         NA      NA       NA    
AgriblockPanruti                             NA         NA      NA       NA    
AgriblockSankarapuram                        NA         NA      NA       NA    
Varietydelux ponni                     109.1218    45.9933   2.373   0.0177 *  
Varietyponmani                        -147.4697    45.4718  -3.243   0.0012 ** 
`Soil Types`clay                        11.8758    36.0021   0.330   0.7415    
Nurserywet                              37.8201    36.0241   1.050   0.2939    
`Wind Direction_D1_D30`ENE                   NA         NA      NA       NA    
`Wind Direction_D1_D30`NW                    NA         NA      NA       NA    
`Wind Direction_D1_D30`SSE                   NA         NA      NA       NA    
`Wind Direction_D1_D30`SW                    NA         NA      NA       NA    
`Wind Direction_D1_D30`W                     NA         NA      NA       NA    
`Wind Direction_D31_D60`NE                   NA         NA      NA       NA    
`Wind Direction_D31_D60`S                    NA         NA      NA       NA    
`Wind Direction_D31_D60`W                    NA         NA      NA       NA    
`Wind Direction_D31_D60`WNW                  NA         NA      NA       NA    
`Wind Direction_D61_D90`NNE                  NA         NA      NA       NA    
`Wind Direction_D61_D90`NNW                  NA         NA      NA       NA    
`Wind Direction_D61_D90`SE                   NA         NA      NA       NA    
`Wind Direction_D61_D90`SW                   NA         NA      NA       NA    
`Wind Direction_D91_D120`NW                  NA         NA      NA       NA    
`Wind Direction_D91_D120`S                   NA         NA      NA       NA    
`Wind Direction_D91_D120`SSE                 NA         NA      NA       NA    
`Wind Direction_D91_D120`W                   NA         NA      NA       NA    
`Wind Direction_D91_D120`WSW                 NA         NA      NA       NA    
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Residual standard error: 944.6 on 2778 degrees of freedom
Multiple R-squared:  0.9895,	Adjusted R-squared:  0.9895 
F-statistic: 2.617e+04 on 10 and 2778 DF,  p-value: < 2.2e-16

```{r}
# vif of model 1 
library(car)
vif(model_1)
```
```{r}
alias(model_1)
```
Independent Variables
 `Seedrate(in Kg)` `30DRain( in mm)` `30_50DRain( in mm)` `Min temp_D1_D30` `Max temp_D1_D30` `Min temp_D31_D60` Varietydelux ponni Varietyponmani `Soil Types`clay Nurserywet
 
 

## Model 2
```{r}
# model with only independent variables

feature_set_2 <- c("Seedrate(in Kg)", "30DRain( in mm)", "30_50DRain( in mm)", "Min temp_D1_D30", "Max temp_D1_D30", "Min temp_D31_D60", "Variety", "Soil Types", "Nursery")
data_2 <- data_drop_cor %>%
  select(all_of(feature_set_2), `Paddy yield(in Kg)`)
model_2 <- lm(`Paddy yield(in Kg)` ~ ., data = data_2)
summary(model_2)
plot(model_2)


```








