---
title: "newdata"
author: "Gerardo Gutierrez"
date: "2025-12-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load and Inspect Data
```{r}
# Load necessary libraries
library(dplyr) # data wrangling
library(tidyr) # data tidying and reshaping
library(ggplot2) # pretty plots
library(readr) 
# Load the dataset
raw_df <- read_csv("../data/agricultural_yield_train.csv")
```


```{r}
raw_df
```

```{r}
str(raw_df)
```

```{r}
colnames(raw_df)
```
[1] "Soil_Quality"                     "Seed_Variety"                     "Fertilizer_Amount_kg_per_hectare" "Sunny_Days"                      
[5] "Rainfall_mm"                      "Irrigation_Schedule"              "Yield_kg_per_hectare"            

```{r}
# Check for missing values
colSums(is.na(raw_df))
```




```{r}
# correlation heatmap for numeric var
library(reshape2)
numeric_df <- raw_df %>% select_if(is.numeric)
cor_matrix <- cor(numeric_df)
melted_cor_matrix <- melt(cor_matrix)
ggplot(data = melted_cor_matrix, aes(x=Var1, y=Var2, fill=value)) + 
  geom_tile() +
  geom_text(aes(label=round(value,2)), color="white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       limit = c(-1,1), name="Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                   size = 12, hjust = 1)) +
  coord_fixed()
```
No strong correlation among covariates.


```{r}
# unique values of seed variety
unique(raw_df$Seed_Variety)

```



```{r}
# irrigation schedule unique values
unique(raw_df$Irrigation_Schedule)
```

```{r}
# histogram of all variables 
library(gridExtra)
p1 <- ggplot(raw_df, aes(x=Fertilizer_Amount_kg_per_hectare)) +
  geom_histogram(binwidth=10, fill="blue", color="black", alpha=0.7) +
  labs(title="Distribution of Fertilizer Amount", x="Fertilizer Amount (kg/ha)", y="Count") +
  theme_minimal()
p2 <- ggplot(raw_df, aes(x=Sunny_Days)) + 
  geom_histogram(binwidth=1, fill="blue", color="black", alpha=0.7) +
  labs(title="Distribution of Sunny Days", x="Sunny Days", y="Count") +
  theme_minimal()
p3 <- ggplot(raw_df, aes(x=Rainfall_mm)) +
  geom_histogram(binwidth=10, fill="blue", color="black", alpha=0.7) +
  labs(title="Distribution of Rainfall", x="Rainfall (mm)", y="Count") +
  theme_minimal()
p4 <- ggplot(raw_df, aes(x=Irrigation_Schedule)) +
  geom_histogram(binwidth=1, fill="blue", color="black", alpha=0.7) +
  labs(title="Distribution of Irrigation Schedule", x="Irrigation Schedule", y="Count") +
  theme_minimal()
p5 <- ggplot(raw_df, aes(x=Seed_Variety)) +
  geom_histogram(binwidth=1, fill="blue", color="black", alpha=0.7) +
  labs(title="Distribution of Seed Variety", x="Seed Variety", y="Count") +
  theme_minimal()
p6 <- ggplot(raw_df, aes(x=Soil_Quality)) +
  geom_histogram(binwidth=1, fill="blue", color="black", alpha=0.7) +
  labs(title="Distribution of Soil Quality", x="Soil Quality", y="Count") +
  theme_minimal()
p7 <- ggplot(raw_df, aes(x=Yield_kg_per_hectare)) +
  geom_histogram(binwidth=10, fill="blue", color="black", alpha=0.7) +
  labs(title="Distribution of Yield", x="Yield (kg/ha)", y="Count") +
  theme_minimal()
grid.arrange(p1, p2, p3, p4, p5, p6, p7, ncol=3)

```






# Model Selection

## model 1 - full model 
```{r}
# model with yiled as target
model1 <- lm(Yield_kg_per_hectare ~ Soil_Quality + Seed_Variety + Fertilizer_Amount_kg_per_hectare + Sunny_Days + Rainfall_mm + Irrigation_Schedule, data = raw_df)
summary(model1)
plot(model1)
```



```{r}
# suppress warnings
options(warn=-1)
# Durbin-Watson test for autocorrelation
library(lmtest)
dwtest(model1)

```
```{r}
# verify there is no multicollinearity with VIF
library(car)
vif(model1)
```
VIF values are all close to 1, furhter suggesting no multicollinearity.



## model 2 - without irrigation schedule
```{r}

# model without irrigation schedule
model2 <- update(model1, . ~ . - Irrigation_Schedule)
summary(model2)
plot(model2)
```
R- Squared severely dropped by about .3, - suggesting that we should keep irrigation schedule in the model.

## model 3 - without seed variety
```{r}
# model without seed variety
model3 <- update(model1, . ~ . - Seed_Variety)
summary(model3)
plot(model3)
```
Notice that the QQplot now shows severe violation of normality. Suggesting that seed variety is importatn for statistical tests. 
Additionally, the R-squared also dropped by about .5, suggesting that seed variety is important for model fit.

## model 4 - without sunny days
```{r}
# model without sunny days
model4 <- update(model1, . ~ . - Sunny_Days)
summary(model4)
plot(model4)
```
R-squared only dropped by about .02 when sunny days was removed, suggesting that sunny days may not be very important for model fit but lets conduct anova test to confirm that sunny days is not significant. 
```{r}
# compare reduced model with full model using ANOVA to see if sunny days is significant
anova(model1, model4)
```
$H_0:$ Sunny days does not significantly improve the model fit.
We REJECT $H_0$ since p-value is less than 0.05, suggesting that sunny days is significant and should be kept in the model.

## model 5 - without fertilizer amount
```{r}
# model without fertilizer amount
model5 <- update(model1, . ~ . - Fertilizer_Amount_kg_per_hectare)
summary(model5)
plot(model5)
```
```{r}
# compare reduced model with full model using ANOVA to see if fertilizer amount is significant
anova(model1, model5)
```
Ferlizer helpls variability significantly.
R-squared dropped by about .1 when fertilizer amount was removed.

Additionally, in the anova test we reject $H_0:$ fertilizer amount does not significantly improve the model. So we keep fertilizer in the model.

## model 6 - without rainfall
```{r}
# model without rainfall
model6 <- update(model1, . ~ . - Rainfall_mm)
summary(model6)
plot(model6)
```
```{r}
# compare reduced model with full model using ANOVA to see if rainfall is significant
anova(model1, model6)
```
We keep rainfall since it helps explain variabality and is significant in the anova F test.



```{r}
# compare models using AIC
AIC(model1, model2, model3, model4, model5, model6)

```
In conclusion, we keep the full model (model 1) as it has the lowest AIC, highest adjusted R-squared, and all variables are significant based on anova tests.

# Inference and Interpretations
**Interpretation of beta coefficients:**

```{r}
summary(model1)$coefficients
```

- Holding all other covariates constant, we can expect an average increase in yield of about 1.54 kg/ha for each additional unit increase in soil quality.
- Switching to a high yield seed variety (from 0 to 1), we can expect an average increase in yield of about 300.46 kg/ha, holding all other variables constant.
- For each additional kg/ha of fertilizer applied, we can expect an average increase in yield of about 0.81 kg/ha, holding all other variables constant.
- For each additional sunny day, we can expect an average increase in yield of about 1.99 kg/ha, holding all other variables constant.
- For each additional mm of rainfall, we can expect an average decrease in yield of about 0.51 kg/ha, holding all other variables constant.
- For each addition in irrigation count, we can expect an average increase in yield of about 49.99 kg/ha, holding all other variables constant.

## investigate relationship between rainfall and irrigation
```{r}
ggplot(raw_df, aes(x=Rainfall_mm, y=Irrigation_Schedule)) +
  geom_point(color="blue", alpha=0.6) +
  geom_smooth(method="lm", color="red") +
  labs(title="Relationship between Rainfall and Irrigation Schedule", x="Rainfall (mm)", y="Irrigation Schedule") +
  theme_minimal()
```
```{r}
# investigate why excessive rainfall is negatively impacting yield
cor(raw_df$Rainfall_mm, raw_df$Irrigation_Schedule)

```


## check for outliers, influential, and leverage points


```{r}
# Leverage points
n = nobs(model1)
p = length(coef(model1))

leverages <- hatvalues(model1)

# amount of high leverage points
high_lev_points <- which(leverages > (3 * p/n))
length(high_lev_points)
# fitted values of high leverage points
model1$fitted.values[leverages > (3 * p/n)]

# fit model without high leverage points
df_no_high_lev <- raw_df[-high_lev_points, ]
model_no_high_lev <- lm(Yield_kg_per_hectare ~ ., data = df_no_high_lev)
summary(model_no_high_lev)
# compare coefficients
model1$coefficients
model_no_high_lev$coefficients
# compare r-squared
summary(model1)$r.squared
summary(model_no_high_lev)$r.squared
```

```{r}
# Outliers
standardized_residuals <- rstandard(model1)
studentized_residuals <- rstudent(model1)

model1$fitted.values[abs(studentized_residuals) > 3]
model1$fitted.values[abs(standardized_residuals) > 3]
```
```{r}
# remove all outliers
outlier_rows <- which(abs(standardized_residuals) > 3)
outlier_rows

# fit model without outliers
df_no_outliers <- raw_df[-outlier_rows, ]
model_no_outliers <- lm(Yield_kg_per_hectare ~ ., data = df_no_outliers)
summary(model_no_outliers)

# compare coefficients
model1$coefficients
model_no_outliers$coefficients

# compare r -squared
summary(model1)$r.squared
summary(model_no_outliers)$r.squared


```
Model is robust to outliers - coefficients and R-squared do not change significantly after removing outliers.
```{r}
# influential
cook_distances <- cooks.distance(model1)
model1$fitted.values[cook_distances > 1]
```
```{r}
# dffits
n = nobs(model1)
p = length(coef(model1))

dffits_values <- dffits(model1)
model1$fitted.values[abs(dffits_values) > (2 * sqrt(p/n))]

# remove influential points based on dffits
influential_dffits_rows <- which(abs(dffits_values) > (2 * sqrt(p/n)))
influential_dffits_rows
# fit model without influential dffits points
df_no_influential_dffits <- raw_df[-influential_dffits_rows, ]
model_no_influential_dffits <- lm(Yield_kg_per_hectare ~ ., data = df_no_influential_dffits)
summary(model_no_influential_dffits)
```
```{r}
# Dfbetas
n = nobs(model1)
p = length(coef(model1))

dfbetas_values <- data.frame(dfbetas(model1))
abs(dfbetas_values) > (2 / sqrt(n))
```

```{r}
# remove influential points where there is at least 1 dfbeta
dfbeta_rows_1 <- which(rowSums(abs(dfbetas_values) > (2 / sqrt(n))) >= 1)
# size of influential_rows
length(dfbeta_rows_1)

# build model without influential observations
df_no_dfbeta_1 <- raw_df[-dfbeta_rows_1, ]
model_no_dfbeta_1 <- lm(Yield_kg_per_hectare ~ ., data = df_no_dfbeta_1)
summary(model_no_dfbeta_1)

# compare coefficients
model1$coefficients
model_no_dfbeta_1$coefficients

# compare r squared
summary(model1)$r.squared
summary(model_no_dfbeta_1)$r.squared
```

```{r}
# remove influential points where there are at least 4 dfbetas
dfbeta_rows_4 <- which(rowSums(abs(dfbetas_values) > (2 / sqrt(n))) >= 4)
# size of influential_rows
length(dfbeta_rows_4)
# build model without influential observations
df_no_dfbeta_4 <- raw_df[-dfbeta_rows_4, ]
model_no_dfbeta_4 <- lm(Yield_kg_per_hectare ~ ., data = df_no_dfbeta_4)
summary(model_no_dfbeta_4)
# compare coefficients
model1$coefficients
model_no_dfbeta_4$coefficients
# compare r squared
summary(model1)$r.squared
summary(model_no_dfbeta_4)$r.squared

```


After removing high dfbeta observations, regression coefficients are not being heavily influenced by the points with high dfbetas.
However, R-squared increased slightly after removing all high DFBETA points so they may be affecting model fit slightly.

```{r}
# 
confint(model1, level = 0.95)
```

**Interpretation of confidence intervals for beta coefficients:**
- We are 95% confident that, holding all other variables constant, each unit increase in soil quality will result in between 1.49 and 1.60 kg/ha of additional yield.
- We are 95% confident that switching to a high yield seed variety will result in between 298.77 and 302.16 kg/ha of additional yield, holding all other variables constant.
- We are 95% confident that for each additional kg/ha of fertilizer applied, the yield will increase by between 0.80 and 0.82 kg/ha, holding all other variables constant.
- We are 95% confident that for each additional sunny day, the yield will increase by between 1.91 and 2.07 kg/ha, holding all other variables constant.
- We are 95% confident that for each additional mm of rainfall, the yield will decrease by between 0.51 and 0.50 kg/ha, holding all other variables constant.
- We are 95% confident that for each additional irrigation count, the yield will increase by between 49.64 and 50.33 kg/ha, holding all other variables constant.




# Prediction Intervals and Confidence Interval
```{r}
# preidction interval for new data
new_farm <- data.frame(
  Soil_Quality = 80,
  Seed_Variety = 1,
  Fertilizer_Amount_kg_per_hectare = 200,
  Sunny_Days = 100,
  Rainfall_mm = 500,
  Irrigation_Schedule = 8
)

predict(model1, newdata = new_farm, interval = "prediction")

```


```{r}
# confidence intervals for mean prediction
predict(model1, newdata = new_farm, interval = "confidence")

```

```{r}
# visualize prediction and confidence intervals
library(ggplot2)
predictions <- predict(model1, interval = "prediction")
conf_intervals <- predict(model1, interval = "confidence")
pred_df <- data.frame(
  Fitted = model1$fitted.values,
  Lower_Pred = predictions[, "lwr"],
  Upper_Pred = predictions[, "upr"],
  Lower_Conf = conf_intervals[, "lwr"],
  Upper_Conf = conf_intervals[, "upr"],
  Actual = raw_df$Yield_kg_per_hectare
)
ggplot(pred_df, aes(x=Fitted, y=Actual)) +
  geom_point(color="blue", alpha=0.6) +
  geom_ribbon(aes(ymin=Lower_Pred, ymax=Upper_Pred), fill="red", alpha=0.2) +
  geom_ribbon(aes(ymin=Lower_Conf, ymax=Upper_Conf), fill="green", alpha=0.2) +
  labs(title="Prediction and Confidence Intervals", x="Fitted Values", y="Actual Yield") +
  theme_minimal()


```

The prediction band shows that most of the true yields fall within the 95% prediction interval for new observations and that the model predicts new data well.


#Prediction 
We will test predictive power of model by splitting data into train test sets and using prediction metrics such as RMSE and MAE, MSPE.

```{r}
# split data into train and test sets
set.seed(123)
train_indices <- sample(1:nrow(raw_df), size = 0.7 * nrow(raw_df))
train_df <- raw_df[train_indices, ]
test_df <- raw_df[-train_indices, ]
# fit model on train data
model_train <- lm(Yield_kg_per_hectare ~ ., data = train_df)
# predict on test data
predictions_test <- predict(model_train, newdata = test_df)
# calculate RMSE, MAE, MSPE
rmse <- sqrt(mean((test_df$Yield_kg_per_hectare - predictions_test)^2))
mae <- mean(abs(test_df$Yield_kg_per_hectare - predictions_test))
mspe <- mean((test_df$Yield_kg_per_hectare - predictions_test)^2)

# cacluate average percent error 
ape <- mean(abs((test_df$Yield_kg_per_hectare - predictions_test) / test_df$Yield_kg_per_hectare)) * 100

cat("RMSE:", rmse, "\n")
cat("MAE:", mae, "\n")
cat("MSPE:", mspe, "\n")
cat("Average Percent Error:", ape, "%\n")

```
```{r}
# Plot distribution of yield data to compare with prediction errors
ggplot(raw_df, aes(x=Yield_kg_per_hectare)) +
  geom_histogram(binwidth=10, fill="blue", color="black", alpha=0.7) +
  labs(title="Distribution of Yield Data", x="Yield (kg/ha)", y="Count") +
  theme_minimal()
```

```{r}
# visualize predicted vs actual
library(ggplot2)
ggplot(data = data.frame(Actual = test_df$Yield_kg_per_hectare, Predicted = predictions_test), aes(x=Actual, y=Predicted)) +
  geom_point(color="light blue", alpha=0.6) +
  geom_abline(slope=1, intercept=0, color="red", linetype="dashed") +
  labs(title="Actual vs Predicted Yields", x="Actual Yield", y="Predicted Yield") +
  theme_minimal()
```

Confidence bands show how precise the estimates are. Where as the predictive power shows how well it generalizes to unseen data. 

# Recommendations
Based on the analysis, the following recommendations can be made to farmers aiming to maximize their agricultural yield:
1. **Soil Quality Improvement:** Invest in soil enhancement techniques such as adding organic matter, proper pH balancing, and nutrient management to improve soil quality, as it has a positive impact on yield.
2. **Seed Variety Selection:** Opt for high-yield seed varieties, as they significantly increase yield compared to low-yield varieties.
3. **Fertilizer Application:** Apply an optimal amount of fertilizer (around 200 kg/ha) to boost yield, while avoiding over-application which may not lead to proportional yield increases.
4. **Maximize Sunny Days Exposure:** Plan planting schedules to take advantage of periods with higher sunny days, as they contribute positively to yield.
5. **Rainfall Management:** Implement effective water management strategies to mitigate the negative impact of excessive rainfall on yield, such as proper drainage systems. However, we should also consider that this factor may require further investigation as to why it negatively impacts yield. As consistent rainfall is generally beneficial for crops but maybe more spread out rainfall is better than heavy rainfall in short periods.
6. **Irrigation Scheduling:** Maintain an effective irrigation schedule (around 8 times) to ensure crops receive adequate water, which positively influences yield.


